The Overlapping Control Volume Finite Element Method for Multiphase Porous Media Flow Modelling \\ Part I: Numerical Formulation

### 2015-01-21 Papers/Prototype_1_Method/article.tex
Paper 1 is updated taking into account comments from Pain, Percival and Jackson. Different from the previous version of the paper (under SVN), results of the Buckley-Leverett problems in 1-3D were introduced (based on existing plots and comments of Prototype Paper 2

\begin{abstract} 
This is the first of two papers devoted to a novel finite element method formulation to solve subsurface multi-fluid flows. The overlapping control volume finite element method (OCVFEM) has two major strengths, firstly it uses a dual consistent pressure-velocity representation in control volume and finite element spaces which ensures local mass conservation. Secondly, it uses a family of finite element types that ensures multiphase Darcy equations can be exactly enforced. The underlying mass conservation equations are solved in control volume space and finite elements are used to obtain high-order fluxes on control volume boundaries. The novelty of the method lies in (a) permitting both continuous and discontinuous description of pressures and saturations between elements; (b) the use of arbitarily high-order polynomial representation for pressure and velocity within the mixed formulation; (c) the use of high-order flux-limited methods to avoid introducing non-physical oscillations as well as achieving high-order time accuracy where and when possible. In this paper, the new model is described and test cases are introduced to show the numerical accuracy in one- and two-dimensional simulations. More extensive applications are introduced in the second part of this two part paper. 
\end{abstract}

@@

Finite difference methods (FDM) have been extensively used in modelling and simulation of fluid flows in porous media~\citep{aziz_1986, chen_1997, chen_2005} however, they are strongly dependent on mesh quality and orientation and cannot easily represent complex geometries. In addition, FDM produce excessive numerical dispersion in heterogeneous porous matrix \citep{chavent_1986}. The geometrical flexibility of finite element methods (FEM) has proven to overcome these deficiencies. Among FEM-based formulations for porous media, the control volume finite element method (CVFEM) has become increasingly popular as it can guarantee local mass conservation and high-order numerical accuracy as well as being able to use tetrahedral geometry-conforming elements~\citep{forsyth_1990, cordazzo_2004, geiger_2004, hurtado_2007}.  \citet{huber_2000} demonstrated that the vertex-centred finite volume - finite element method (or Box scheme) can achieve similar goals~\citep[see also][]{helmig_1997}.

@@

In this paper, we describe a novel overlapping CVFEM (or OCVFEM) discretisation which is conservative and consistent \citep{jackson_2013}. The continuity equations are embedded into the pressure equation to enforce mass conservation and exact force balance. This is possible by using dual CV and FEM discretisations for the pressure equation, while saturation is discretised in CV space. The overall formulation employs an implicit algorithm with respect to time that is unconditionally stable for pure advection and with no restriction on the time-step size.

The formulation has two key developments. Firstly, it uses a dual consistent pressure-velocity representation in CV and FEM spaces embedded in a family of triangular and tetrahedral finite element-pairs, P$_{n}$DG-P$_{m}$[DG] \citep[see][]{cotter_2009b}. In this element type, the dual velocity and pressure fields are represented by $n^{\rm th}$ order (discontinuous) and $m^{\rm th}$ order (continuous/discontinuous) polynomials, respectively. This allows the velocity to exactly represent pressure gradients in the flow solution for homogeneous material properties. The P$_{n+1}$DG-P$_{n}$DG element-pair, which has similar properties to the P$_n$DG-P$_{n+1}$ element-pair, allows a representation in which pressure, %\jrp{Saturation is on the control volume space and thus always discontinuous} 
saturation and other solution variables are all fully discontinuous across finite element boundaries. The second key development is that this formulation uses an overlapping finite element representation of velocity within each element. This approach can achieve an exact representation of the Darcy force balance equations.

\medskip
This is the first of two papers on the proposed methodology. Here we focus on model description and initial validation, whereas in the second part~\citep{pavlidis_2013} we employ the numerical formulation for a number of multiphase porous media flow simulations to demonstrate numerical accuracy and convergence rates. The remainder of this paper is organised as follows. New element types and the overlapping formulation are introduced in Sections~\ref{element_types_section} and~\ref{overlapping_method_section}.  A detailed description of the new porous media methods is given in Section~\ref{transport_methods_section}. Scalar field equations, upwinding schemes and solution methods are introduced in Sections~\ref{Section:Saturation_Global}-\ref{Section:SolvingLinearEqns}. Initial model validation is presented in Section~\ref{Section:Results} and final remarks are drawn in Section~\ref{conc}.

%%%         %%%
%%% SECTION %%%
%%%         %%%
\section{Continuous and Discontinuous Finite Element Types and Control Volumes} \label{element_types_section}
The family of P$_{n}$DG-P$_{n+1}$ element-pairs was originally introduced by \citet{cotter_2009a} for geophysical fluid dynamics applications \citep[see also][]{cotter_2012}. In particular, the P$_{1}$DG-P$_{2}$ element-pair was originally developed to represent the balance of geostrophic pressure and velocity without introducing spurious pressure modes. This element-pair uses element-wise linear polynomial FE basis functions, discontinuous across element interfaces for velocity $\left(\text{P}_{1}\text{DG}\right)$, and quadratic polynomial finite element basis functions for pressure which maintain $C_{\text{0}}$ continuity across element boundaries $\left(\text{P}_{2}\right)$. %This is analogous to the exact balance also achieved with this element between the frictional and pressure gradient terms in Darcy's equations.

Figure~\ref{fem_cv_represent_a} displays three element types of the P$_{n}$DG-P$_{n+1}$ family that are used in this set of papers. P$_{\text{0}}$DGP$_{\text{1}}$ and P$_{1}$DG-P$_{2}$ are shown in Fig.~\ref{fem_cv_represent_a}(a) and (b), respectively, where pressure nodes are stored in the vertices and edges (in P$_{\text{2}}$ elements) of triangular/tetrahedral elements and control volumes are constructed around them based on a Voronoi decomposition.  One can realise that due to the continuous finite element node structure, control volumes span more than one element. Having CV's spanning to multiple elements may create artificial diffusion in cases where there are large variations in physical properties across elements, such as at the boundary between materials with different absolute permeabilities. As an alternative, two families of fully discontinuous element-pairs are also considered: P$_{n}$DG-P$_{n}$DG and P$_{n+1}$DG-P$_{n}$DG~\citep[Fig.~\ref{fem_cv_represent_a}(c), applications using these element-pairs are exploited in][]{pavlidis_2015} along with control volumes from their dual Voronoi decompositions. These families of element-pairs are LBB stable and free from chequerboard pressure modes, as previously demonstrated by~\citet{cotter_2009b}~\citep[see also][]{cotter_2011}.

@@

%%%         %%%
%%% SECTION %%%
%%%         %%%
\section{Overlapping-CVFEM Formulation -- An Accurate Representation of Multiphase Darcy Flows}
\label{overlapping_method_section}
Darcy's law for immiscible multiphase flow may be written \citep{chen_2006} in the form:%,bear_1972}:
\begin{equation}\label{e:darcy_eqn}
  \mathbf{q}_{\alpha} = -\frac{\mathcal{K}_{{r}_\alpha}\mathbf{K}}{\mu_{\alpha}}\left( \nabla p_{\alpha} - {\mathbf{s}_{u}}_{\alpha} \right),
\end{equation}
where $\mathbf{q}_{\alpha}=\left(q_\alpha^{(1)},q_\alpha^{(2)},q_\alpha^{(3)}\right)^{\text{T}}$ is the $\alpha$-th phase Darcy flow rate for 3D flows, $\mathbf{K}$ is the absolute permeability tensor of the porous matrix, $\mathcal{K}_{{r}_\alpha}\left(S_{\alpha}\right)$ is the phase relative permeability, which is a function of the phase saturation $S_{\alpha}\left(\mathbf{r},t\right)$. $\mu_{\alpha}$, $p_{\alpha}$ and $\mathbf{s}_{{u}_\alpha}$ are the phase dynamic viscosity, pressure and a source term, respectively.

If we define the interstitial or advective phase velocity, averaged over the entire medium, as $\mathbf{u}_\alpha= \mathbf{q}_\alpha/S_\alpha$, then we may rewrite Eqn.~\ref{e:darcy_eqn} as:
\begin{equation}
  \bm{v}_\alpha:= {\underline {\underline \sigma}}_{\alpha} \mathbf{u}_{\alpha} = - \nabla p + {\mathbf{s}_{u}}_{\alpha},
  \label{force-bal}
\end{equation}
where ${\underline {\underline \sigma}}_{\alpha}=\mu_\alpha S_\alpha \left(\mathcal{K}_{{r}_\alpha}\mathbf{K}\right)^{-1}$ represents the implicit linearisation of the viscous frictional forces and $\bm{v}_\alpha$ is a saturation averaged vector of the force density of the phase, related to the advective velocity through $\bm{u}_\alpha={\underline {\underline \sigma}}_{\alpha}^{-1}\bm{v}_\alpha$.% arising in the derivation of extended Darcy's law.

In order to discretise Eqn.~\ref{force-bal}, we assume a finite element representation for the force and pressure fields and a control volume representation for phase saturation fields. The representations for force density and pressure in terms of their FE basis functions $Q_{j}$ and $P_{j}$  may be expressed, respectively, as:
\begin{displaymath}
  \bm{v}_\alpha(\bm{r},t) = \sum\limits_{j=1}^{\mathcal{N}_u} Q_{j}(\bm{r})\bm{v}_{\alpha,j}(t) \;\;\;\;\text{ and } \;\;\;\; p(\bm{r},t)  = \sum\limits_{j=1}^{\mathcal{N}_p} P_{j}(\bm{r})p_{j}(t).
\end{displaymath} 
Here $\mathcal{N}_{u}$ and $\mathcal{N}_{p}$ are the total number of degrees of freedom for the FE force and pressure representations. Each component of the weak form of the force balance (Eqn.~\ref{force-bal}) is tested with the force density basis function space to obtain:
\begin{eqnarray}
  \sum\limits_{E} \left. \int\limits_{\Omega_E} { {Q}}_i
   \left ( {\mathbf v}_\alpha + \nabla p
  -{\mathbf s}_{u_\alpha} \right) dV \right. + \displaystyle
  \oint_{\Gamma_{E}} {Q}_i {\mathbf n} \left(p - \tilde{p}\right) d\Gamma
  + \nonumber \\ 
  \oint_{\Gamma_{\Omega}} { Q}_i {\mathbf n} \left(p -
  p_{bc}\right) d\Gamma = \bm{0},
  \label{force-semi-disc} 
\end{eqnarray} 
where $\Omega_E$ and $\Gamma_{E}$ are the volume and boundary of element $E$, respectively, and $\Gamma_{\Omega}$ is the boundary of the computational domain. The numerical pressure appearing in the jump condition, $\tilde{p}$, is an average of the potentially discontinuous pressure across the element $E$,
\[\tilde{p}=\lim_{a\rightarrow 0} 
   \frac{1}{2}\left[p(\bm{r}+a\bm{n})+p(\bm{r}-a\bm{n})\right].\] This term therefore vanishes when pressure is continuous. Equation ~\ref{force-semi-disc} can be represented, in matrix form, as:
\begin{equation}
  {\mathbf M} \underline {\mathbf v} = -{\mathbf C} \underline {\mathbf p} + \underline {\mathbf s}_{u}, \label{force-balance-matrix-form}
\end{equation}
where the velocity $\left(\underline{\bf v}\right)$ and pressure $\left(\underline{\bf p}\right)$ solution-vectors are defined as,
\begin{eqnarray}
  &&\underline{{\bf v}} = \left({v_x}_{1,1},\ldots,{v_x}_{\mathcal{N}_\alpha,
    \mathcal{N}_u}, {v_y}_{1,1},\ldots,{v_y}_{\mathcal{N}_\alpha,
    \mathcal{N}_u},{v_z}_{1,1},\ldots,{v_z}_{\mathcal{N}_\alpha,
    \mathcal{N}_u}\right)^{T} \;\;\;\;\text{and} \nonumber \\ &&
  \underline{{\bf p}} = \left(p_1, p_2, p_3, ...,
  p_{\mathcal{N}_p}\right)^T. \nonumber \nonumber
\end{eqnarray}
$\mathcal{N}_{\alpha}$ is the number of phases and $v_x$, $v_y$, $v_z$ are the the velocity components associated with the $x-$ and $y-$ and $z-$ coordinates, respectively.

Finally, the mass matrix ${\mathbf M}$, gradient matrix
${\mathbf C}$ and discretised source term ${\mathbf s}_{u}$ are:
\[\mathbf{M}=\left(
\begin{array}{ccc}
  \bm{M}^{(1,1)} &\bm{0} &\bm{0} \\
  \bm{0}   &\bm{M}^{(2,2)} &\bm{0} \\
  \bm{0} &\bm{0}  &\bm{M}^{(3,3)}   
\end{array}\right),\]
\[
\bm{C}=\left(
\begin{array}{ccc}
  \bm{C}^{(1,1)}&\bm{0}&\bm{0}\\
  \bm{0}&\bm{C}^{(2,2)}&\bm{0}\\
  \bm{0}&\bm{0}&\bm{C}^{(3,3)}\\
\end{array}\right)\quad \text{ and } \;\;
\bm{s}_u=\left(
\begin{array}{ccc}
  \bm{s}^{(1,1)}&\bm{0}&\bm{0}\\
  \bm{0}&\bm{s}^{(2,2)}&\bm{0}\\
  \bm{0}&\bm{0}&\bm{s}^{(3,3)}\\
\end{array}\right),\quad
\]
with blocks defined by:
\begin{eqnarray}
  && \left[{\mathbf M}^{(i,j)}\right]_{p,q} = \int_{\Omega}  Q_{p} \delta_{ij}
{Q}_{q}  dV, \\
  && {\left[\bm{C}^{(i,i)}\right]_{p,q} = \int_{\Omega} Q_{p} \frac{\partial P_{q}}{\partial r_i} dV - \oint_{\Gamma_{E}} \displaystyle\frac{1}{2} Q_{p} n_i^{(E,q)} P_{q} d\Gamma -  \oint_{\Gamma_{\Omega}} \displaystyle Q_{p} n_i P_{q} d\Gamma} \\
  && {\left[\bm{s}_{u}^{(i,i)}\right]_{p} = \int_{\Omega} Q_p} \left[{\mathbf s}_{u}\right]_i dV - \oint_{\Gamma_{\Omega}}\displaystyle Q_{p} n_i p_{bc} d\Gamma.
\end{eqnarray}
Here $\delta_{i,j} = \left[\underline{\underline{\sigma}}_{\alpha}\right]_{i,j}$,  $\bm{n}=\left(n_{1},n_{2},n_{3}\right)^{T}$ is the outward-pointing normal vector of the domain $\Omega$. $\bm{n}^{(E,q)}$ is the normal to the boundary of element $E$ and is outward-pointing if $P_{q}$ takes support on $E$, and inward-pointing otherwise.

@@

%This formulation ensures that for uniform \jrp{This tensor needs a good name -- JG: I can't think on a good name for this -- it's too non-linear to hold any physical meaning. Maybe solid-fluid homogeneisation factor/ parameter/ coeff} ${\underline{\underline \sigma}}_{\alpha}$ within each element the balance in the \jrp{is this only true in 1D} 1D extended Darcy flow equations is strictly enforced. 
The saturations are computed in CV space, whereas the permeability tensor $\left(\mathbf{K}\right)$ is assumed piecewise constant in FE space to allow efficient representation of surface parametrised geometries. Thus the viscous-friction damping tensor, \jrp{This tensor needs a good name JG: I can't think on a good name for this -- it's too non-linear to hold any physical meaning. Maybe solid-fluid homogeneisation factor/ parameter/ coeff or viscous-friction damping ...} %${\underline{\underline \sigma}}_{\alpha}$:
\begin{eqnarray}
  {\underline {\underline \sigma}}_{\alpha} &=& {\underline {\underline \sigma}}_{\alpha} \left[S_{\alpha}\left(\mathbf{r},t\right), \mu_{\alpha}, \mathcal{K}_{{r}_\alpha}\left(S_{\alpha}\right), \mathbf{K}\right] \nonumber \\
&=& \frac{S_{\alpha}\mu_{\alpha}}{\mathcal{K}_{{r}_\alpha}}\mathbf{K}^{-1},
\end{eqnarray}
is piecewise constant within the sub-volumes defined through the non-empty intersections between the elements and control volumes; in the pressure discontinuous formulation, these are precisely the control volumes themselves, Fig.~\ref{fem_cv_represent_a}(c). The velocity, formed by the action of ${\underline{\underline \sigma}}_{\alpha}^{-1}$ upon the force density $\bm{v}_\alpha$, has an exact polynomial representation on the sub-volumes of one degree lower that of $p$. For example, for a quadratic pressure variation there will be a piecewise linear velocity field, with discontinuities at element and control volume boundaries (Fig.~\ref{fig:darcys_law}). Thus, in 1D there are $2\times 3$ velocity nodal values per element associated with each phase.

@@

For two neighbouring CV's $i$ and $j$ which have a common FE interface, ${\bf u}_{\alpha}$ can be calculated by using a volume weighting. However, the volume $\mathcal{V}$ of the control volume and $\underline{\underline{\sigma}}_{\alpha}$ act on the velocity ${\bf u}_{\alpha}$ in a very similar way. Hence, the velocity at the interface can be expressed as,
\begin{equation} 
  {\tilde{\bf u}_\alpha} = \frc{ \mathcal{V}_{j} {\bf v}_{\alpha i} + \mathcal{V}_{i} {\bf v}_{\alpha j}} { \mathcal{V}_{i} \underline{\underline{\sigma}}_{\alpha i} + \mathcal{V}_{j} \underline{\underline{\sigma}}_{\alpha j} }.
\end{equation}

@@

\medskip
Similarly, boundary conditions can be applied using Riemann variables to determine if information is travelling into or out of the domain (although most flow simulation models do not follow this more rigorous approach). Specified pressure boundary conditions for outlet flux $\left(\right.$when ${\bf n}\cdot {\bf u}_\alpha>0\left.\right)$ are ${\bf n}\cdot {\bf u}_\alpha {S_\alpha}$, in which neither ${\bf u}_\alpha$ nor ${S_\alpha}$ are specified.

>>>>

\begin{abstract} 
This is the first of two papers devoted to a novel finite element method formulation to solve subsurface multi-fluid flows. The overlapping control volume finite element method (OCVFEM) has two major strengths, firstly it uses a dual consistent pressure-velocity representation in control volume and finite element spaces which ensures local mass conservation. Secondly, it uses a family of finite element types that ensures multiphase Darcy equations can be exactly enforced. The underlying mass conservation equations are solved in control volume space whereas finite elements are used to obtain high-order fluxes on control volume boundaries. The novelty of the method lies in (a) permitting both continuous and discontinuous description of pressures and saturations between elements; (b) the use of arbitarily high-order polynomial representation for pressure and velocity within the mixed formulation and; (c) the use of high-order flux-limited methods to avoid introducing non-physical oscillations as well as achieving high-order time accuracy where and when possible. In this paper, the new model is described and test cases are introduced to show the numerical accuracy in one- and two-dimensional simulations. More extensive applications are introduced in the second part of this two-part paper. 
\end{abstract}

@@

Finite difference methods (FDM) have been extensively used in modelling and simulation of fluid flows in porous media~\citep{aziz_1986, chen_1997, chen_2005} however, they are strongly dependent on mesh quality and orientation and cannot easily represent complex geometries. In addition, FDM produce excessive numerical dispersion in heterogeneous porous matrix \citep{chavent_1986}. The geometrical flexibility of finite element methods (FEM) has proven to overcome these deficiencies. Among FEM-based formulations for porous media, the control volume finite element method (CVFEM) has become increasingly popular as it can guarantee local mass conservation and high-order numerical accuracy as well as being able to use tetrahedral geometry-conforming elements~\citep{forsyth_1990, cordazzo_2004, geiger_2004, hurtado_2007}.  \citet{huber_2000} demonstrated that the vertex-centred finite volume -- finite element method (or Box scheme) can achieve similar goals~\citep[see also][]{helmig_1997}.

@@

In this paper, we describe a novel overlapping CVFEM (or OCVFEM) formulation which is conservative and consistent. The continuity equations are embedded into the pressure equation to enforce mass conservation and exact force balance. This is possible by using dual CV and FEM discretisations for the pressure equation, while saturation is discretised in CV space. The overall formulation employs an implicit algorithm with respect to time that is unconditionally stable for pure advection and with no restriction on the time-step size.

The formulation has two key developments. Firstly, it uses a dual consistent pressure-velocity representation in CV and FEM spaces embedded in a family of triangular and tetrahedral finite element-pairs, \PN[n]{m}[DG]~\citep[see][]{cotter_2009b}. In this element type, the dual velocity and pressure fields are represented by $n^{\rm th}$ order (discontinuous) and $m^{\rm th}$ order (continuous/discontinuous) polynomials, respectively. Such flexibility allows the velocity to exactly represent pressure gradients in the flow solution for homogeneous material properties (e.g., \PN[1]{2}). The \PNDG[n+1]{n} element-pair, which has similar properties to the \PN[n]{n+1} element-pair, allows a representation in which pressure, saturation and other solution variables are all fully discontinuous across finite element boundaries. The second key development is that this formulation uses an overlapping finite element representation of velocity within each element. This approach can achieve an exact representation of the Darcy force balance equations.

\medskip
This is the first of two papers on the proposed methodology. Here we focus on model description and initial validation, whereas in the second part, \citet{pavlidis_2015} apply the numerical formulation to a number of multiphase porous media flow problems to demonstrate numerical accuracy and convergence rates. The remainder of this paper is organised as follows. New element types and the overlapping formulation are introduced in Sections~\ref{element_types_section} and~\ref{overlapping_method_section}.  Detailed description of the new methods is given in Section~\ref{transport_methods_section}-\ref{Section:SolvingLinearEqns}. Initial model validation is presented in Section~\ref{Section:Results} and final remarks are drawn in Section~\ref{conc}.

%%%         %%%
%%% SECTION %%%
%%%         %%%
\section{Continuous and Discontinuous Finite Element Types and Control Volumes} \label{element_types_section}
The family of P$_{n}$DG-P$_{n+1}$ element-pairs was originally introduced by \citet{cotter_2009a} for geophysical fluid dynamics applications \citep[see also][]{cotter_2012}. In particular, the P$_{1}$DG-P$_{2}$ element-pair was developed to represent the balance of geostrophic pressure and velocity without introducing spurious pressure modes. This element-pair uses element-wise linear polynomial FE basis functions, discontinuous across element interfaces for velocity $\left(\text{P}_{1}\text{DG}\right)$, and quadratic polynomial finite element basis functions for pressure which maintain $C_{\text{0}}$ continuity across element boundaries $\left(\text{P}_{2}\right)$. %This is analogous to the exact balance also achieved with this element between the frictional and pressure gradient terms in Darcy's equations.

Figure~\ref{fem_cv_represent_a} displays three element types of the P$_{n}$DG-P$_{n+1}$ family that are used in this set of papers. P$_{\text{0}}$DGP$_{\text{1}}$ and P$_{1}$DG-P$_{2}$ are shown in Fig.~\ref{fem_cv_represent_a}(a) and (b), respectively, where pressure nodes are stored in the vertices and edges (in P$_{\text{2}}$ elements) of triangular/tetrahedral elements. Control volumes are constructed around them based on a Voronoi decomposition.  One can realise that due to the continuous finite element node structure, control volumes span more than one element. Having CV's spanning to multiple elements may create artificial diffusion in cases where there are large variations in physical properties across elements, such as at the boundary between materials with different absolute permeabilities. As an alternative, two families of fully discontinuous element-pairs are also considered: \PNDG[n]{n} and \PNDG[n+1]{n}~\citep[Fig.~\ref{fem_cv_represent_a}(c), applications using these element-pairs are exploited in][]{pavlidis_2015} along with control volumes from their dual Voronoi decompositions. These families of element-pairs are LBB stable and free from chequerboard pressure modes, as previously demonstrated by~\citet{cotter_2009b}~\citep[see also][]{cotter_2011}.

@@

%%%         %%%
%%% SECTION %%%
%%%         %%%
\section{Overlapping-CVFEM Formulation -- An Accurate Representation of Multiphase Darcy Flows}
\label{overlapping_method_section}
Darcy's law for immiscible multiphase flow may be written \citep{chen_2006} in the form:%,bear_1972}:
\begin{equation}\label{e:darcy_eqn}
  \mathbf{q}_{\alpha} = -\frac{\mathcal{K}_{{r}_\alpha}\mathbf{K}}{\mu_{\alpha}}\left( \nabla p_{\alpha} - {\mathbf{s}_{u}}_{\alpha} \right),
\end{equation}
where $\mathbf{q}_{\alpha}=\left(q_\alpha^{(1)},q_\alpha^{(2)},q_\alpha^{(3)}\right)^{\text{T}}$ is the $\alpha$-th phase Darcy flow rate for 3D flows, $\mathbf{K}$ is the absolute permeability tensor of the porous matrix, $\mathcal{K}_{{r}_\alpha}\left(S_{\alpha}\right)$ is the phase relative permeability, which is a function of the phase saturation $S_{\alpha}\left(\mathbf{r},t\right)$. $\mu_{\alpha}$, $p_{\alpha}$ and $\mathbf{s}_{{u}_\alpha}$ are the phase dynamic viscosity, pressure and source term, respectively.

If we define the interstitial or advective phase velocity, averaged over the entire medium, as $\mathbf{u}_\alpha= \mathbf{q}_\alpha/S_\alpha$, then we may rewrite Eqn.~\ref{e:darcy_eqn} as:
\begin{equation}
  \mathbf{v}_{\alpha}:= {\underline {\underline \sigma}}_{\alpha} \mathbf{u}_{\alpha} = - \nabla p + {\mathbf{s}_{u}}_{\alpha},
  \label{force-bal}
\end{equation}
where ${\underline {\underline \sigma}}_{\alpha}=\mu_\alpha S_\alpha \left(\mathcal{K}_{{r}_\alpha}\mathbf{K}\right)^{-1}$ represents the implicit linearisation of the viscous frictional forces and $\mathbf{v}_\alpha$ is a saturation averaged vector of the force density of the phase, related to the advective velocity through $\mathbf{u}_\alpha={\underline {\underline \sigma}}_{\alpha}^{-1}\mathbf{v}_\alpha$.% arising in the derivation of extended Darcy's law.

In order to discretise Eqn.~\ref{force-bal}, we assume a finite element representation for the force and pressure fields and a control volume representation for phase saturation fields. The representations for force density and pressure in terms of their FE basis functions $Q_{j}$ and $P_{j}$  may be expressed, respectively, as:
\begin{displaymath}
  \mathbf{v}_\alpha(\bm{r},t) = \sum\limits_{j=1}^{\mathcal{N}_u} Q_{j}(\bm{r})\mathbf{v}_{\alpha,j}(t) \;\;\;\;\text{ and } \;\;\;\; p(\bm{r},t)  = \sum\limits_{j=1}^{\mathcal{N}_p} P_{j}(\bm{r})p_{j}(t).
\end{displaymath} 
Here $\mathcal{N}_{u}$ and $\mathcal{N}_{p}$ are the total number of degrees of freedom for the FE force and pressure representations. Each component of the weak form of the force balance (Eqn.~\ref{force-bal}) is tested with the force density basis function space to obtain:
\begin{eqnarray}
  \sum\limits_{E} \left. \int\limits_{\Omega_E} { {Q}}_i
   \left ( {\mathbf v}_\alpha + \nabla p
  -{\mathbf s}_{u_\alpha} \right) dV \right. + \displaystyle
  \oint_{\Gamma_{E}} {Q}_i {\mathbf n} \left(p - \tilde{p}\right) d\Gamma
  + \nonumber \\ 
  \oint_{\Gamma_{\Omega}} { Q}_i {\mathbf n} \left(p -
  p_{bc}\right) d\Gamma = \bm{0},
  \label{force-semi-disc} 
\end{eqnarray} 
where $\Omega_E$ and $\Gamma_{E}$ are the volume and boundary of element $E$, respectively, and $\Gamma_{\Omega}$ is the boundary of the computational domain. The numerical pressure appearing in the jump condition, $\tilde{p}$, is an average of the potentially discontinuous pressure across the element $E$,
\[\tilde{p}=\lim_{a\rightarrow 0} 
   \frac{1}{2}\left[p(\bm{r}+a\bm{n})+p(\bm{r}-a\bm{n})\right].\] This term therefore vanishes when pressure is continuous. Equation ~\ref{force-semi-disc} can be represented, in matrix form, as:
\begin{equation}
  {\mathbf M} \underline {\mathbf v} = -{\mathbf C} \underline {\mathbf p} + \underline {\mathbf s}_{u}, \label{force-balance-matrix-form}
\end{equation}
where $\underline{\bf v}$ and $\underline{\bf p}$ solution-vectors are defined as,
\begin{eqnarray}
  &&\underline{{\bf v}} = \left({v_x}_{1,1},\ldots,{v_x}_{\mathcal{N}_\alpha,
    \mathcal{N}_u}, {v_y}_{1,1},\ldots,{v_y}_{\mathcal{N}_\alpha,
    \mathcal{N}_u},{v_z}_{1,1},\ldots,{v_z}_{\mathcal{N}_\alpha,
    \mathcal{N}_u}\right)^{T} \;\;\;\;\text{and} \nonumber \\ &&
  \underline{{\bf p}} = \left(p_1, p_2, p_3, ...,
  p_{\mathcal{N}_p}\right)^T. \nonumber \nonumber
\end{eqnarray}
$\mathcal{N}_{\alpha}$ is the number of phases and $v_x$, $v_y$, $v_z$ are the components associated with the $x-$ and $y-$ and $z-$ coordinates, respectively.

Finally, the mass matrix ${\mathbf M}$, gradient matrix
${\mathbf C}$ and discretised source term ${\mathbf s}_{u}$ are:
\[\mathbf{M}=\left(
\begin{array}{ccc}
  \bm{M}^{(1,1)} &\bm{0} &\bm{0} \\
  \bm{0}   &\bm{M}^{(2,2)} &\bm{0} \\
  \bm{0} &\bm{0}  &\bm{M}^{(3,3)}   
\end{array}\right),\]
\[
\bm{C}=\left(
\begin{array}{ccc}
  \bm{C}^{(1,1)}&\bm{0}&\bm{0}\\
  \bm{0}&\bm{C}^{(2,2)}&\bm{0}\\
  \bm{0}&\bm{0}&\bm{C}^{(3,3)}\\
\end{array}\right)\quad \text{ and } \;\;
\bm{s}_u=\left(
\begin{array}{ccc}
  \bm{s}^{(1,1)}&\bm{0}&\bm{0}\\
  \bm{0}&\bm{s}^{(2,2)}&\bm{0}\\
  \bm{0}&\bm{0}&\bm{s}^{(3,3)}\\
\end{array}\right),\quad
\]
with blocks defined by:
\begin{eqnarray}
  && \left[{\mathbf M}^{(i,j)}\right]_{p,q} = \int_{\Omega}  Q_{p} \delta_{ij}
{Q}_{q}  dV, \nonumber \\
  && {\left[\bm{C}^{(i,i)}\right]_{p,q} = \int_{\Omega} Q_{p} \frac{\partial P_{q}}{\partial r_i} dV - \oint_{\Gamma_{E}} \displaystyle\frac{1}{2} Q_{p} n_i^{(E,q)} P_{q} d\Gamma -  \oint_{\Gamma_{\Omega}} \displaystyle Q_{p} n_i P_{q} d\Gamma}, \nonumber \\
  && {\left[\bm{s}_{u}^{(i,i)}\right]_{p} = \int_{\Omega} Q_p} \left[{\mathbf s}_{u}\right]_i dV - \oint_{\Gamma_{\Omega}}\displaystyle Q_{p} n_i p_{bc} d\Gamma. \nonumber
\end{eqnarray}
Here $\delta_{i,j} = \left[\underline{\underline{\sigma}}_{\alpha}\right]_{i,j}$,  $\bm{n}=\left(n_{1},n_{2},n_{3}\right)^{T}$ is the outward-pointing normal vector of the domain $\Omega$. $\bm{n}^{(E,q)}$ is the normal to the boundary of element $E$ and is outward-pointing if $P_{q}$ takes support on $E$, and inward-pointing otherwise.

@@

%This formulation ensures that for uniform \jrp{This tensor needs a good name -- JG: I can't think on a good name for this -- it's too non-linear to hold any physical meaning. Maybe solid-fluid homogeneisation factor/ parameter/ coeff} ${\underline{\underline \sigma}}_{\alpha}$ within each element the balance in the \jrp{is this only true in 1D} 1D extended Darcy flow equations is strictly enforced. 
The saturations are computed in CV space, whereas the permeability tensor $\left(\mathbf{K}\right)$ is assumed piecewise constant in FE space to allow efficient representation of surface parametrised geometries. Thus the viscous-friction damping tensor, %\jrp{This tensor needs a good name JG: I can't think on a good name for this -- it's too non-linear to hold any physical meaning. Maybe solid-fluid homogeneisation factor/ parameter/ coeff or viscous-friction damping ...} %${\underline{\underline \sigma}}_{\alpha}$:
\begin{eqnarray}
  {\underline {\underline \sigma}}_{\alpha} &=& {\underline {\underline \sigma}}_{\alpha} \left[S_{\alpha}\left(\mathbf{r},t\right), \mu_{\alpha}, \mathcal{K}_{{r}_\alpha}\left(S_{\alpha}\right), \mathbf{K}\right] \nonumber \\
&=& \frac{S_{\alpha}\mu_{\alpha}}{\mathcal{K}_{{r}_\alpha}}\mathbf{K}^{-1},
\end{eqnarray}
is piecewise constant within the sub-volumes defined through the non-empty intersections between the elements and control volumes; in the pressure discontinuous formulation, these are precisely the control volumes themselves, Fig.~\ref{fem_cv_represent_a}(c). The velocity, formed by the action of ${\underline{\underline \sigma}}_{\alpha}^{-1}$ upon the force density $\mathbf{v}_\alpha$, has an exact polynomial representation on the sub-volumes of one degree lower that of $p$. For example, for a quadratic pressure variation there will be a piecewise linear velocity field, with discontinuities at element and control volume boundaries (Fig.~\ref{fig:darcys_law}). Thus, in 1D there are $2\times 3$ velocity nodal values per element associated with each phase.

@@

For two neighbouring CV's $i$ and $j$ with common FE interface, ${\bf u}_{\alpha}$ can be calculated by using a volume weighting. However, the volume $\mathcal{V}$ of the control volume and $\underline{\underline{\sigma}}_{\alpha}$ act on the velocity ${\bf u}_{\alpha}$ in a very similar way. Hence, the velocity at the interface can be expressed as,
\begin{equation} 
  {\tilde{\bf u}_\alpha} = \frc{ \mathcal{V}_{j} {\bf v}_{\alpha i} + \mathcal{V}_{i} {\bf v}_{\alpha j}} { \mathcal{V}_{i} \underline{\underline{\sigma}}_{\alpha i} + \mathcal{V}_{j} \underline{\underline{\sigma}}_{\alpha j} }.
\end{equation}

@@

Similarly, boundary conditions can be applied using Riemann variables to determine if information is travelling into or out of the domain (although most flow simulation models do not follow this more rigorous approach).
